Towards a Practical Face Recognition System: \\ Robust Alignment and Illumination by Sparse Representation

### 2010-12-25 rebuttal.tex

\documentclass[11pt]{article}

\usepackage{epsfig}
\usepackage{latexsym, amsmath, amsfonts}
%\usepackage{pdfsync}

\renewcommand{\baselinestretch}{1.0}
\parskip 2.2mm
\parindent 0mm
\topmargin -0.60in \oddsidemargin 0.0625in \textheight 9.00in
\textwidth 6.50in

\renewcommand{\Re}{{\mathbb R}}
\newcommand{\Ze}{{\mathbb Z}}
\def\QED{~\rule[-1pt]{5pt}{5pt}\par\medskip}
\newenvironment{proof}{{\bf Proof: \ }}{ \hfill \QED}

\newcommand{\ie}{{\it i.e., }}
\newcommand{\D}{\displaystyle}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}
\newtheorem{assumption}{Assumption}
\newtheorem{definition}{Definition}

\begin{document}

   \bibliographystyle{plain}

   \title{\Large {\bf Manuscript  Revision Statement}}

       \author{{Andrew Wagner, John Wright, Arvind Ganesh, Zihan Zhou, Hossein Mobahi, and Yi Ma}}

   \date{}
   \maketitle

We thank the reviewers for their constructive comments, and in particular
reviewer 1 for an extra in-depth review.  While we have philosophical
differences on several points, we deeply appreciate that the valuable constructive 
criticisms raised by the reviewers have helped us strengthen the paper considerably.  
Below we will respond to comments of the reviewers in order as well as indicate 
specific changes that we have made to address their concerns.

\section{Respond to Reviewer \#1}
Reviewer 1 makes several specific recommendations for improving the
presentation of the paper, many of which we have heeded.  Please refer to the change
list for details.

\paragraph{About Presentation:} The reviewer has suggested we can reduce the description
of the ALM algorithm in Section II-D. We agree ALM has been extensively studied in the optimization 
literature. We do not claim any novelty here. However, we believe a complete description of the algorithm 
here is important for the TPAMI audience: There are
many algorithms for solving the associated optimization problem, with drastically different complexity
and speed. Typical vision researchers might not be so familiar with some of the recent
developments in these algorithms. In particular some of the more effective improvements to ALM are fairly 
recent and scattered in different literatures. In fact, many of the questions raised by reviewer \#2
ask for even more detailed explanation about the algorithms. So we believe, a complete but brief 
description of the ALM-based algorithm is important for the audience of TPAMI to implement and 
test the ideas efficiently and correctly.

We find other suggestions on presentation useful and have made changes accordingly.

\paragraph{About Positioning and Novelty:}
As we have tried to make it clear in the very beginning, the proposed method does not aim to solve face 
recognition for all scenarios! It is not designed for the case where the training is uncontrolled and limited, hence it
may not have any advantages over other face recognition techniques/systems specifically designed for such scenarios. 
However, there are many practical scenarios where the proposed setting is very relevant: access control
or recognizing refugees, where cooperative training images can (and should) be obtained in advance but no cooperative test
images are needed or available. No need of cooperative test images is arguably the main (or only) advantage that 
face recognition has over other biometric measures for security, such as fingerprint, iris, and DNA. 
If the subjects can cooperate for both training and testing, as suggested by the reviewers, then face 
recognition may no longer be the best choice, in terms of cost and accuracy. The proposed system is an attempt
to build a system that is not only accurate and robust enough to compete with other biometric measures but also 
discreet and non-intrusive when in use. To the best of our knowledge (probably the reviewer may agree), even 
for this admittedly restricted scenario, there is no practical face-recognition system whose performance can meet
the standards of most access control and security purposes. Existing systems have found success  mostly in applications
that are less-demanding in performance such as online image search or photo album organization etc.

\paragraph{About Summary Recommendation:} While we agree with the reviewer that careful evaluation of our
system is needed by comparing it with other face recognition techniques/systems. We must point out that most face recognition systems are not designed for the scenario considered by this paper, hence a fair and meaningful comparison can be difficult. Nevertheless, to better evaluate both the strength and limitation of our method, we have added experimental comparison of our system with LBP, one of the leading techniques used for face verification for most commercial face recognition systems (in addition to other techniques already compared in the paper). We compare both methods with the same settings on two datasets that are representative of controlled and uncontrolled training: the Multi-PIE and the FERET
datasets.  The results indicate both suboptimal performance of LBP for applications where many controlled training images (under different illuminations) are available, and conversely, suboptimal performance of our algorithm on the FERET dataset where the training is not controlled nor sufficient.

\paragraph{Detailed Concerns or Comments by Reviewer 1:}
\begin{enumerate}
\item As we have mentioned earlier, we completely agree with the reviewer that if the users are completely cooperative at
both training and testing phases, face recognition may not necessarily be the best choice of biometric measure -- fingerprint
and iris are arguably better choices. However, our method does NOT require the users to be cooperative at the test
stage. The purpose of having a good training set and a robust recognition system is precisely to alleviate the users 
from being so cooperative at the test stage. This will allow the system to be less intrusive and much more user-friendly 
as it can handle typically daily change of faces such as occlusion by glasses, hats, or beard and variation in lighting and shadow. Ideally, face recognition will take place automatically and unknowingly as a user walks through a door or works in front a desktop.

In addition, if the accuracy and reliability of a face recognition system can truly get close to that of fingerprint and iris, it can serve as a complementary measure in a multi-modal system and improve the overall security for high-stake applications. To the best of our knowledge, no existing face recognition system, with or without cooperative training, comes close to meet the standards of high-stake security applications.

Furthermore, we have found that the use of the L1-norm as a robust error function improves performance 
even when the test image does not contain occlusions.  Any boundaries where the intensity changes in a
misaligned image can result in a large localised error component.  While
unoccluded faces are mostly smooth, sharp boundaries still occur where there
are cast shadows, especially at depth discontinuities such as the nostrils, the
sides of the nose, and partially open mouths.

\item Reviewer 1's interest in face recognition for terrorist watchlist and law
enforcement applications is of limited relevance to the evaluation of this
article.  By claiming that ``the field of face recognition has set itself the
goal of recognition with as few gallery images as possible,'' the reviewer is
conflating the constraints of one particular type of applications of face recognition
(surveillance) with those of face recognition as a whole. This is a (noble) goal probably easy to
set than to meet. Although progresses have been made towards this goal, the reviewer
probably has to agree that after several decades (and several high-profile failures of
face recognition systems in real-world testing), people are still looking for techniques 
that can achieve the performance required by high-end applications such as homeland 
security. We believe that, before such techniques can be found or such a goal can be
achieved at all, we should not preventing ourselves from making progress in developing systems that 
can already provide good guarantee for success, even if the conditions are somewhat limited.

A huge part of the message of our paper is that for applications that allow it, we can and should
be gathering more and higher quality data at enrollment to our advantage. In fact, majority 
of the crime watch-list consists of criminals who have been committed before. Today,
the acquisition standards for fingerprints are far more strict than that for
capturing face images of criminals and foreigners -- precisely because they are shown to
help fingerprint identification systems. Of course, the idea of better training helps is certainly
not new or even recent.  What is new is the confluence of: 1. The availability
of hardware to efficiently acquire such data (high intensity, high contrast
ratio DLP projectors), 2. Robust and efficient computational tools to use such data in a
principled fashion, 3. Computer hardware fast enough to even contemplate using
entire images as features.

\item ``It is conceivable that SRC is applicable to any feature as well - which
underscores the fact that it is simply a pattern classification toolbox - on
par with any found in a textbook on pattern recognition. What is it that makes
it special for face recognition?''

Most general pattern classification algorithms do not take into account the
physically motivated linear model that is central (though by no means unique)
to the problem of illumination variation.

\item Our goal is not to achieve the best recognition rates on the NIST family of
benchmarks (FERET, FRVT, FRGC, MBGC, GBU); rather it is to advance state of
the art in face recognition given cooperative subjects.  The motivating
application of the NIST data sets is government law enforcement (originally for
enforcement of drug prohibition, and more recently shifting towards border
security).  This requires backwards compatibility with old image data, some of
which was gathered before computers and digital cameras were even in widespread use.  
Due to these historical reasons, none of these data sets contain multiple controlled
illuminations.  For example, FRGC contains at most two illuminations per person
in the gallery.  (Two studio lights were kept on, and a third light was
switched either on or off.) The techniques used to gather these data sets are not a wise 
choice for applications where users are cooperative (foreigners applying for a visa)
or forced to be so (suspects or criminals in custody), and are needlessly
restrictive.  While large public data sets are critical to the rigorous
comparison of recognition algorithms, their existence should not be used to
argue against the development of new data acquisition techniques and algorithms
appropriate for them.

\item We have added comparison with a Local Binary Patterns based recognition system, and
have shown that both our system and LBP perform better in the application domains or settings
for which they were designed.  Please see the change list for details.

\item Although we claim to have a preliminary system that works in a realistic
access control setting, we have never claimed that our method solves all problems in face
recognition. We do not know where the reviewer gets that wrong impression. 
From the very beginning and then at multiple places later, we actually emphasize
the importance of having the more restrictive training data for guaranteeing the success 
of our method.

If it helps, we will try to tone down the rhetoric in XYZ places. UNFINISHED!!

\item We will add further discussion contrasting the advantages and disadvantages of our
system vs. systems incorporating 3D models. This paper does not intend to deal with large 3D pose
although the proposed techniques are directly applicable to the cases when images of more poses or 
even a full 3D models are available.

\item The issues of (mis)alignment and illumination variance are coupled in the
sense that if one problem is solved, the other problem becomes a lot easier. But solving them
separately is not sufficient for solving the face recognition problem.

\item If the data fits our model perfectly, at the correct alignment the only
corruptions are due to occlusions.  Until the images are perfectly aligned,
there is indeed a component of dense error caused by misalignment.  This dense
error component is indeed largest where the image gradient is high (or even
discontinuous), and the derivative of this error component is what drives the
algorithm to its solution.  We explicitly do blur and downsample the images to
mollify registration effects; multiscale alignment is a well-known technique,
and is not new to our paper.

\item We have removed the reference to scattered light, as there appear to
be several common uses of the term.  That said, it is true that linear illumination
models are physically motivated for scenes far more general than for convex 
lambertian objects.  Maxwell's equations in free space are linear, and materials
that create non-linear boundary conditions are the exception
rather than the rule.  Without proportionality between radiance and irradiance 
the concept of a Bidirectional Reflectance Distribution Function and its
generalizations are not even meaningful.

\item Rigorous justification of the choice of illuminations is a subject of 
continuing research.  Our current choice is motivated empirically by the
experiments described in the paper.

\item To our knowlege, we are the first to use projectors for illumination in
face recognition applications.  Section 1A of the previous submission already
cites several other applications that use projectors as illumination sources.
The requested references propose a technique for reducing the SNR of projector
based illumination sysems for certain classes of objects.  We have added
discussion of this to the paper.

\item  It is true that one or both stages of the algorithm could be substituted
with other techniques.  That said, the registration and recognition stages are
coherent in the sense that they are motivated by same modeling assumptions
(illumination model, sparse occlusion, etc.), they use the same robust penalty
on image error (the L1-norm), and they use the same numerical optimization
techniques (ALM).  While conceptual simplicity is a secondary concern in
algorithm design, starting with a minimalist foundation benefits both
practitioners as well as researchers.  Other researchers have already published
more complex extensions to our work.

\item Since we perform alignment separately for each class in the gallery, the
computation for the alignment stage is linear in the size of the gallery.  The
variation in alignment across classes is indeed small, but still changes the
objective enough to affect recognition results.  Aligning to a generic face (or
a gallery based on ``aligned'' images of many people) is a good idea, and one we
have considered as well.  If nothing else, it could be used to improve
alignment initialization instead of using raw face detector output.  It may
also be useful to use the roughly aligned test image to quicky eliminate some
of the classes from further consideration.  Again, we have deliberately tried
to keep the algorithm as minimalist as possible for this publication.

\item The adaptation of the Lucas-Kanade algorithm to both the L1 norm cost
function and a linear illumination model is novel to our knowlege.  As
discussed earlier, we disagree with Reviewer 1's contention that data
acquisition is not a critical part of the face recognition problem -- careful (training) 
data acquisition has been a critical part for almost all other biometric identification 
systems (e.g. fingerprint and iris and DNA)! Why face recognition needs to be so different?

We believe we have demonstrated that using a well-though-out combination of existing ideas
(iterative image alignment, L1-error function, SRC, using projectors for
illumination), one can build a recognition system that is conceptually simple, well
motivated, and competitive with state-of-the-art recognition systems for access
control scenarios.

\end{enumerate}

\section{Respond to Reviewer \#2}
Reviewer 2 raises 5 questions, which we will now address.  Since the
recognition optimization is convex and has no difficultly converging, we assume
that (1)-(3) are asking about the iterative alignment scheme.

\begin{enumerate}
\item (i) ``How often an infinite loop is encountered, and how it is handled."

We place a hard limit on the maximal number of alignment iterations, so the
algorithm is guaranteed to terminate in finite time.  In practice, the
alignment algorithm tends to converge within 4-10 iterations for each
resolution of the multi-scale implementation.  We also place hard bounds on how
large the alignment can be, and terminate alignment early if these bounds are
reached.  In the cases where the alignment ``slips off'' the face, this is how
alignment terminates.

\item (ii) ``How often the starting point happens to be located in a wrong basin of
attraction, i.e. a wrong face. (If there is any way to know this.)''

Since we align the test image to the training users individually, we avoid the
potential problem of the test image converging to a local minimum corresponding
to a non-match face.

\item (iii) ``How often the solution gets drifted away into a wrong basin of
attraction. (If there is any way to know this.)''

In Section II E, we have conducted careful experiments to measure the region of attraction
for the proposed alignment method.

\item (iv) ``What is the criteria or threshold for convergence? If the threshold is
stringent, the algorithms are likely to get stuck in infinite loops, and if the
threshold is liberal, premature convergence would occur. How are such
thresholds picked and how universal are they? Does every new dataset need its
own threshold?''

For the l1 minimizations we threshold the l1-norm of the difference between
successive values of the state (this includes the both the explicit
optimization variables and the Largrange multipliers).  Similarly, for the
alignment iteration, we threshold the l1-norm of the update to the
transformation parameters.  We used the same set convergence thresholds for all
of the data sets we ran against.  With the suggested thresholds, the l1-minimzations are
guaranteed to converge eventually, although convergence may be very slow if they
are chosen poorly.

\item (v) ``The problem, however, is that if we minimize a function one variable at
a time, we may actually end up increasing the function $(L_{\mu}(x,e,\lambda_{k}))$ value.''

This may be true for general nonlinear optimization problem. However for convex problems
we are solving here, the alternating direction methods have been proven to converge, see
[27] and related work by the same authors.

\end{enumerate}

\section{Respond to Reviewer \#3}
Reviewer 3 makes the following requests:

\begin{enumerate}
\item ``I would have liked to see the experiment for the case S=1, without
impostors in the dataset, to judge the effect of the ``Sparse
Concentration Index''.

We have added experiments to the paper demonstrating the utility of the
sci-based recognition stage for both identification and imposter rejection.  In
general, we see a few percentage points of improvement in recognition rate when
using SCI over recognition based purely on alignment residual.  
It is worth emphasizing that while this effect is desirable, 
the primary value of SCI is its use for imposter rejection.

\item ``I would also retitle the paper to include ``robust measure''
instead of ``sparse representation", as the L1 Cost is just a robust
cost, and the system would probably work as well or better with other
(even non-convex) robust cost functions.''

We hope reviewer will find our supplemental justification for the (Sparse
Representation based) recognition convincing.  If so, the reviewer may
therefore also agree that the current title is indeed appropriate.

\item ``It might be possible to streamline the introduction a bit more, such that it
clearly describes the algorithm before going into the details.''

We have attempted to make the overall structure of our system clear earlier in the
paper, by clarifying early on that SRC forms the recognition stage of our system
in the introduction, as well as by making it clear in the caption of the motivating example
is not the final pipeline.

\item ``Using and requiring several input images of a person asks for a
comment on the advantage or difference to a system that uses
a 3D representation of a face. Please comment on this!"

The primary advantage of systems that internally use a 3D model of human faces
is that they have the potential to work over a wider range of pose variation in
the test image.  In general, systems using 2D models are simpler and easier to
engineer.  As far as data acquisition is concerned, 2D cameras still have a far
higher resolution and lower noise levels than 3D cameras.  Based on the principles
of photometric stereos, our set of 2D images may very well encode more useful shape 
information about a user's face than a single low-resolution depth image, especially if 
recognition is to be performed on 2D input images.

\end{enumerate}

\end{document}

### 2010-12-28 faces_system.tex
Changes from Yi, handing off to John.

\PARstart{T}{here} is a historical tendency for face recognition algorithms to work well under laboratory conditions but degrade when tested in less-controlled environments.  While there have been several high profile failed trials of face recognition technology for extremely difficult mass surveillance / watch-list applications, face recognition has not even seen widespread applications in more tractable settings, such as access control for buildings, computer systems, automobiles, or even automatic teller machines.\footnote{Face recognition scenarios where both gallery and test data are uncontrolled is an active research area as well \cite{LFW}.}  These applications are very interesting due to their potential sociological impact.  They are also qualitatively different from surveillance applications because they demand extremely reliable systems, and because the gallery subjects are allies, rather than opponents, of the recognition system.  This creates the possibility of carefully controlling the acquisition of the training data, even if the testing data may be less controlled for usability reasons.

Classical holistic subspace based methods
\cite{Turk1991-CVPR,Belhumeur1997-PAMI} are well known for their speed and
simplicity, as well as for their natural extension to linear illumination
models.  However, their performance has been shown to be extremely brittle not
only to alignment variation, but to even minor occlusions caused by, say, a
wisp of hair, a blinked eye, or mouth that is slightly open. One promising
recent direction, set forth in \cite{Wright2009-PAMI}, casts the recognition
problem as one of finding a sparse representation of the test image in terms of
the training set as a whole, up to some sparse error due to occlusion. A
\emph{sparse representation-based classification} (SRC) method is then proposed
for recognition. The main idea is that the sparse nonzero coefficients should
concentrate on the training samples with the same class label as the test
sample. SRC has demonstrated striking recognition performance despite severe
occlusion or corruption by solving a simple convex program.  For this
reason, the final recognition stage of our proposed system will consist of SRC,
and the alignment stage that precedes it will be based on similar techniques.

@@

Unfortunately, while SRC achieves impressive results, it does not deal with misalignment
between the test and training images, and it requires a rich set of illuminations
in the gallery images for good performance.  
%Furthermore, the experiments were carried out on public datasets taken under
%controlled laboratory conditions such as Extended Yale B
%\cite{Georghiades2001-PAMI}, which do not contain realistic test images.
%two critical aspects of real world face recognition must be addressed:  significant variations in both the {\em image domain} and in
%the {\em image value}.
We illustrate this with an example in Figure \ref{fig:promo}.
The task is to identify the girl among 20 subjects. If the test
face image, say obtained from an off-the-shelf face detector,
has even a small amount of registration error against the
training images (caused by mild pose, scale, or misalignment),
the representation is no longer informative, even if sufficient
illuminations are present in the training, as shown in Figure
\ref{fig:promo}(top). In addition, in order to linearly span
the illuminations of a typical indoor (or outdoor) environment,
illuminations from behind the subject are also needed in the
training. Otherwise, even for perfectly aligned test images,
the representation will not necessarily be sparse or
informative, as shown by the example in Figure
\ref{fig:promo}(middle). Clearly, both good alignment, as well
as sufficient training images are needed.  
%Unfortunately, most public face databases lack images with a
%significant component of rear (more than 90 degrees from
%frontal) illumination, either for training or testing.

@@

We briefly review existing techniques for recognition, image registration, and
handling of illumination variation.  Our system is based purely on 2D
techniques.  This fact immediately distinguishes our technique from systems
that either require a 3D data as an input, or attempt to estimate a 3D model
from input 2D data \cite{amberg2007reconstructing,Blanz2003-PAMI}.  While these
techniques have been shown to achieve better robustness to pose variation given
a sufficiently accurate 3D model, for access control applications where the
user is cooperative in aligning their head to the camera, robustness to
moderate misalignment is sufficient.  Furthermore, many 3D techniques rely on
multiple views to estimate their model, and may have trouble if all of the
training images are captured from one pose.  Note that images of people's faces
under varying illumination contain shape-related discriminative information,
and this information can be leveraged by 2D algorithms even if shape is not
reconstructed explicitly.

% Alignment
In holistic recognition algorithms, correspondence between points in the test
image and in the training must be achieved.  A long line of research exists on
using Active Appearance Models \cite{Cootes2001-PAMI}, and the closely related
Active Shape Models \cite{cootes1992active} to register images against a
relatively high-dimensional model of plausible face appearances, often
leveraging face-specific contours.  While these model-based techniques have
advantages in dealing with variations in expression and pose, they may add
unnecessary complexity to a system where subjects are willing to present a
neutral expression.  We prefer to focus on warpings with far fewer parameters,
and to use the training images themselves as the appearance model. Iterative
registration in this spirit dates at least back to the Lucas-Kanade algorithm
\cite{lucas1981iterative}.

However, whereas much of the early work on image registration is aimed at the
problem of registering nearly identical images, say by minimizing a sum of
squared distances, here we must confront several physical factors
simultaneously: misalignment, illumination variations, and corrupted pixels.
As we discuss further below, illumination variation can be dealt by
expressing the test image as a linear combination of an appropriate set of
training images. Similar representations have been exploited in
illumination-robust tracking (e.g., \cite{Belhumeur1999-PAA,Murase1995-IJCV}).
For robustness to gross errors, the $\ell^1$-norm of the residual is a more
appropriate objective function than the classical $\ell^2$-norm. Its use here
is loosely motivated by theoretical results due to Candes and Tao
\cite{CandesE2005-IT} (see also \cite{Wright2008-IT}). These two observations
lead us to pose the registration problem as the search for a set of
transformations and illumination coefficients that minimize the $\ell^1$-norm
of the representation error. We solve this problem using a generalized
Gauss-Newton method which solves a sequence of affine-constrained $\ell^1$-norm
minimization problems \cite{Osborne1990-JAMSSB,Jittorntrum1980-NM}. Each of
these problems can also be solved efficiently using recently developed
first-order techniques for $\ell^1$-minimization, which are reviewed in
\cite{YangA2010-pp}.

% Illumination
Researchers have tried various techniques to deal with illumination variation.
In almost all recognition algorithms where only a single gallery image is
available per individual, illumination effects are regarded as a nuisance that
must be removed before the algorithm can continue.  This is typically done by
making statistical assumptions about how illumination affects the image, and
using those assumptions to extract a new representation that is claimed to be
illumination invariant.  Recent examples include \cite{chen2006total} and
\cite{zhou2007appearance}.  However, despite these efforts, truly
illumination-invariant features are difficult to obtain from a single input
image.  We argue that if one has the luxury of designinig the recognition algorithm,
but also the acquisition system, and the application demands a high recognition rate,
it is inadvisable to limit the gallery to a
single image per person.  We therefore take the strategy of sampling many
gallery images of each individual under varying illuminations.  These images
are used as the basis for either a convex cone model
\cite{Georghiades2001-PAMI,belhumeur1998set}, or a subspace model
\cite{Basri2003-PAMI}.  Images are captured using a simple-to-construct
projector based light stage.  While similar systems have been used for
other applications, to our knowledge, we
are the first to use projectors to indirectly illuminate a subject's face for
the purpose of face recognition.

\subsection{Contributions} In this paper, we show how the 
mutually confounding issues of registration and illumination
can be naturally addressed within the sparse representation
framework. We show that face registration, a challenging
nonlinear problem, can be solved by a series of linear programs
that iteratively minimize the sparsity of the registration
error. This leads to an efficient and effective alignment
algorithm for face images that works for a large range of
variation in translation, rotation, and scale, even when
the face is only partially visible due to eyeglasses, closed
eyes and open mouth, sensor saturation, etc.  We also propose a
sufficient set of training illuminations
that is capable of interpolating typical indoor and outdoor
lighting, along with a practical hardware system for capturing
them.

@@

\noindent{\bf Organization of this paper.} In Section
\ref{sec:registration}, we derive our robust registration and
recognition algorithm within the sparse representation
framework. We elaborate on the implementation issues of the
algorithm, conduct region of attraction experiments with
respect to both 2D in-plane deformation and 3D pose variation,
and discuss its relationship to existing work in the
literature. Section \ref{sec:illumination} is dedicated to our
training acquisition system. Using this system, we investigate empirically the set of training illuminations required to handle practical illumination variations, and suggest a sufficient set of 38 training illuminations. Extensive experiments on
a large-scale public database and on our own database are conducted
in Section \ref{sec:multipie} and Section \ref{sec:own-data},
respectively, to verify the proposed system. Section
\ref{sec:conclusion} concludes our work with discussion of
promising future directions.

\section{Handling Practical Registration Error}\label{sec:registration}
As demonstrated in Figure \ref{fig:promo}(top), the main
limitation of the {\em sparse representation and
classification} (SRC) algorithm of \cite{Wright2009-PAMI} is
the assumption of pixel-accurate alignment between the test
image and the training set. This leads to brittleness under
pose and misalignment, making it inappropriate for deployment
outside a laboratory setting. In this section, we show how this
weakness can be rectified while still preserving the conceptual
simplicity and good recognition performance of SRC.

>>>>

\PARstart{T}{here} has been a historical tendency for face recognition algorithms to work well under laboratory conditions  (well aligned images, similar lighting conditions, and little change in appearances) but degrade drastically when tested in less-controlled real-world environments. Although achieving high recognition performance under uncontrolled conditions and with limited gallery images has been the traditional goal for face recognition, existing techniques have not yet been able to meet the performance requirements of high-stake security-related applications. While there have been many successes on less-demanding applications such as online image search and family photo album organization (e.g. the freely available Google's Picassa, Microsoft's Photo Gallery, and Apple's iPhoto), there have been many high profile failed trials of face recognition technology worldwide for extremely challenging mass surveillance/watch-list applications.

Somewhat surprisingly, face recognition has not seen widespread applications or success in more tractable security settings, such as access control for secure facilities (e.g., prisons, office buildings), computer systems, automobiles, or even automatic teller machines,  where controlled gallery images can be obtained in advance.\footnote{Face recognition scenarios where both gallery and test data are uncontrolled continue to be an active research area \cite{LFW}, and becomes more popular recently due to aforementioned online applications.}  These applications are very interesting due to their potential sociological and technological impact.  They are also qualitatively different from mass surveillance or online image applications because they demand extremely reliable systems whose performance must compete with other biometric measures such as fingerprint and iris. Now because the gallery subjects are allies, rather than opponents, of the recognition system, this creates the possibility of carefully controlling the acquisition of the training data (as we normally do for all other biometrics). However, unlike other biometrics, face recognition has the potential in working with much less controlled testing data so that the access control system can be made more discreet and much less intrusive than using other biometrics. To some extent, the goal of this paper is to show how a reliable face recognition system can be built for this admittedly restrictive but important scenario.

Classical holistic subspace based face recognition methods \cite{Turk1991-CVPR,Belhumeur1997-PAMI} are well known for their speed and simplicity, as well as for their natural extension to linear illumination
models.  However, their performance has been shown to be extremely brittle not
only to alignment variation, but to even minor occlusions caused by, say, a
wisp of hair, a blinked eye, or mouth that is slightly open. One promising
recent direction, set forth in \cite{Wright2009-PAMI}, casts the recognition
problem as one of finding a sparse representation of the test image in terms of
the training set as a whole, up to some sparse error due to occlusion. A
\emph{sparse representation-based classification} (SRC) method is then proposed
for recognition. The main idea is that the sparse nonzero coefficients should
concentrate on the training samples with the same class label as the test
sample. SRC has demonstrated striking recognition performance despite severe
occlusion or corruption by solving a simple convex program.  
%For this reason, the final recognition stage of our proposed system will consist of SRC,
%and the alignment stage that precedes it will be based on similar techniques.

@@

Unfortunately, while the work \cite{Wright2009-PAMI} achieves impressive results, it does not deal with misalignment
between the test and training images, and it requires a rich set of illuminations
in the gallery images for good performance.  
%Furthermore, the experiments were carried out on public datasets taken under
%controlled laboratory conditions such as Extended Yale B
%\cite{Georghiades2001-PAMI}, which do not contain realistic test images.
%two critical aspects of real world face recognition must be addressed:  significant variations in both the {\em image domain} and in
%the {\em image value}.
We illustrate the compounded effect of both alignment and illumination 
with an example in Figure \ref{fig:promo}.
The task is to identify the girl among 20 subjects. If the test
face image, say obtained from an off-the-shelf face detector,
has even a small amount of registration error against the
training images (caused by mild pose, scale, or misalignment),
the sparse representation obtained using the method of \cite{Wright2009-PAMI} is no longer informative, 
even if sufficient illuminations are present in the training, as shown in Figure
\ref{fig:promo}(top). In addition, in order to span
the illuminations of a typical indoor (or outdoor) environment,
illuminations from behind the subject are also needed in the
training. Otherwise, even for perfectly aligned test images,
the sparse representation obtained using \cite{Wright2009-PAMI} will not necessarily be sparse or
informative, as shown by the example in Figure
\ref{fig:promo}(middle). Clearly, both good alignment, as well
as sufficient training images are needed to ensure success of the sparsity-based 
recognition method proposed by \cite{Wright2009-PAMI}.  
%Unfortunately, most public face databases lack images with a
%significant component of rear (more than 90 degrees from
%frontal) illumination, either for training or testing.
In this paper, we examine how to handle alignment and illumination simultaneously
in the sparse representation framework, hence bring the method proposed in \cite{Wright2009-PAMI}
closer to practical use.

@@

We briefly review existing techniques for recognition, image registration, and
handling of illumination variation.  Our system is based purely on 2D
techniques.  This fact immediately distinguishes our technique from systems
that either require a 3D data as an input, or attempt to estimate a 3D model
from input 2D data \cite{amberg2007reconstructing,Blanz2003-PAMI}.  While these
techniques have been shown to achieve better robustness to pose variation given
a sufficiently accurate 3D model, for access control applications where only 
moderate misalignment and pose are present, the proposed method will be more
than sufficient.   Note that according to the principle of photometric stereo, images of faces
under varying illuminations already contain 3D shape-related information,\footnote{In principle,
one can recover the 3D shape of the face from multiple illuminations using photometric stereo.} 
and this information can be leveraged by 2D algorithms for alignment and recognition 
even if shape is not reconstructed explicitly. In addition, the proposed sparsity-based robust
alignment and recognition techniques are not limited to 2D data and can be easily extended
to 3D data if necessary.

% Alignment
In holistic recognition algorithms, correspondence between points in the test
image and in the training must be achieved.  A long line of research exists on
using Active Appearance Models \cite{Cootes2001-PAMI}, and the closely related
Active Shape Models \cite{cootes1992active} to register images against a
relatively high-dimensional model of plausible face appearances, often
leveraging face-specific contours.  While these model-based techniques have
advantages in dealing with variations in expression and pose, they may add
unnecessary complexity to applications where subjects normally present a neutral face
or only have moderate expression. We prefer to focus on deformations with far fewer 
degrees of freedom (say an affine transform), and to use the training images themselves 
as the appearance model. Iterative registration in this spirit dates at least back to the 
Lucas-Kanade algorithm \cite{lucas1981iterative}.

However, whereas much of the early work on image registration is aimed at the
problem of registering nearly identical images, say by minimizing a sum of
squared distances or maximize normalized correlation, here we must confront several physical factors
simultaneously: misalignment, illumination variations, and corrupted pixels.
As we discuss further below, illumination variation can be dealt by
expressing the test image as a linear combination of an appropriate set of
training images. Similar representations have been exploited in
illumination-robust tracking (e.g., \cite{Belhumeur1999-PAA,Murase1995-IJCV}).
For robustness to gross errors, the $\ell^1$-norm of the residual is a more
appropriate objective function than the classical $\ell^2$-norm. Its use here
is loosely motivated by theoretical results due to Candes and Tao
\cite{CandesE2005-IT} (see also \cite{Wright2008-IT}). These two observations
lead us to pose the registration problem as the search for a set of
transformations and illumination coefficients that minimize the $\ell^1$-norm
of the representation error. We solve this problem using a generalized
Gauss-Newton method which solves a sequence of affine-constrained $\ell^1$-norm
minimization problems \cite{Osborne1990-JAMSSB,Jittorntrum1980-NM}. Each of
these problems can also be solved efficiently using recently developed
first-order techniques for $\ell^1$-minimization, which are reviewed in
\cite{YangA2010-pp}.

% Illumination
Researchers have tried various techniques to deal with illumination variation.
In almost all recognition algorithms where only a single gallery image is
available per individual, illumination effects are regarded as a nuisance that
must be removed before the algorithm can continue.  This is typically done by
making statistical assumptions about how illumination affects the image, and
using those assumptions to extract a new representation that is claimed to be
illumination invariant.  Recent examples include \cite{chen2006total} and
\cite{zhou2007appearance}.  However, despite these efforts, truly
illumination-invariant features are difficult to obtain from a single input
image.  We argue that if one has the luxury of designing the acquisition system 
and the application demands a high recognition rate,
it is then unwise to limit the gallery to a
single image per person.  We therefore take the strategy of sampling many
gallery images of each individual under varying illuminations.  These images
are used as the basis for either a convex cone model
\cite{Georghiades2001-PAMI,belhumeur1998set}, or a subspace model
\cite{Basri2003-PAMI}.  Images are captured using a simple-to-construct
projector based light stage.  While similar systems have been used for
other applications, to our knowledge, we
are the first to use projectors to indirectly illuminate a subject's face for
the purpose of face recognition.

\subsection{Contributions} In this paper, we show how the 
mutually compounding issues of registration and illumination
can be naturally addressed within a robust sparse representation
framework. We show that face registration, a challenging
nonlinear problem, can be solved by a series of linear programs
that iteratively minimize the sparsity of the registration
error. This leads to an efficient and effective alignment
algorithm for face images that works for a large range of
variation in translation, rotation, and scale, even when
the face is only partially visible due to eyeglasses, closed
eyes and open mouth, sensor saturation, etc.  We also propose a
sufficient set of training illuminations
that is capable of interpolating typical indoor and outdoor
lighting, along with a practical hardware system for capturing
them.

@@

\noindent{\em Organization of this paper:} In Section
\ref{sec:registration}, we derive our robust registration and
recognition algorithm within the sparse representation
framework. We elaborate on the implementation issues of the
algorithm, conduct region of attraction experiments with
respect to both 2D in-plane deformation and 3D pose variation,
and discuss its relationship to existing work in the
literature. Section \ref{sec:illumination} is dedicated to our
training acquisition system. Using this system, we investigate empirically the set of training illuminations required to handle practical illumination variations, and suggest a sufficient set of 38 training illuminations. Extensive experiments on
a large-scale public database and on our own database are conducted
in Section \ref{sec:multipie} and Section \ref{sec:own-data},
respectively, to verify the proposed system. Section
\ref{sec:conclusion} concludes our work with discussion of
promising future directions.

\section{Robust Alignment}\label{sec:registration}
As demonstrated in Figure \ref{fig:promo}(top), the main
limitation of the {\em sparse representation and
classification} (SRC) algorithm of \cite{Wright2009-PAMI} is
the assumption of pixel-accurate alignment between the test
image and the training set. This leads to brittleness under
pose and misalignment, making it inappropriate for deployment
outside a laboratory setting. In this section, we show how this
weakness can be rectified while still preserving the conceptual
simplicity and good recognition performance of SRC.
